Introduction
============

The [equation of
motion](https://en.wikipedia.org/wiki/Pendulum_(mathematics)\#Simple_gravity_pendulum)
for a pendulum of unit length subject to [Gaussian white
noise](https://en.wikipedia.org/wiki/White_noise#Mathematical_definitions)
is

$$
\frac{\mathrm{d}^2\alpha}{\mathrm{d}t^2} = -g\sin\alpha + w(t)
$$

We can discretize this via the usual [Euler method](https://en.wikipedia.org/wiki/Euler_method)

$$
\begin{bmatrix}
x_{1,i} \\
x_{2,i}
\end{bmatrix}
=
\begin{bmatrix}
x_{1,i-1} + x_{2,i-1}\Delta t \\
x_{2,i-1} - g\sin x_{1,i-1}\Delta t
\end{bmatrix}
+
\mathbf{q}_{i-1}
$$

where $q_i \sim {\mathcal{N}}(0,Q)$ and

$$
Q
=
\begin{bmatrix}
\frac{q^c \Delta t^3}{3} & \frac{q^c \Delta t^2}{2} \\
\frac{q^c \Delta t^2}{2} & {q^c \Delta t}
\end{bmatrix}
$$

The explanation of the precise form of the covariance matrix will be
the subject of another blog post; for the purpose of exposition of
using Stan and, in particular, Stan's ability to handle ODEs, this
detail is not important.

Instead of assuming that we know $g$ let us take it to be unknown and
that we wish to infer its value using the pendulum as our experimental
apparatus.

[Stan](https://en.wikipedia.org/wiki/Stan_(software)) is a
probabilistic programming language which should be welll suited to
perform such an inference. We use its interface via the R package
[rstan](https://cran.r-project.org/web/packages/rstan/index.html).

A Stan and R Implementation
===========================

Let's generate some samples using Stan but rather than generating
exactly the model we have given above, instead let's solve the
differential equation and then add some noise. Of course this won't
quite give us samples from the model the parameters of which we wish
to estimate but it will allow us to use Stan's ODE solving capability.

Here's the Stan

~~~~{.CPP include="Pendulum.stan"}
~~~~

And here's the R to invoke it

```{r}
library(rstan)
library(mvtnorm)

qc1 = 0.0001
deltaT = 0.01
nSamples = 100
m0 = c(1.6, 0)
g = 9.81
t0 = 0.0
ts = seq(deltaT,nSamples * deltaT,deltaT)

bigQ = matrix(c(qc1 * deltaT^3 / 3, qc1 * deltaT^2 / 2,
                qc1 * deltaT^2 / 2,       qc1 * deltaT
                ),
              nrow = 2,
              ncol = 2,
              byrow = TRUE
              )

samples <- stan(file = 'Pendulum.stan',
                data = list (
                    T  = nSamples,
                    y0 = m0,
                    t0 = t0,
                    ts = ts,
                    theta = array(g, dim = 1),
                    sigma = c(bigQ[1,1], bigQ[2,2]),
                    refresh = -1
                ),
                algorithm="Fixed_param",
                seed = 42,
                chains = 1,
                iter =1
                )
```

We can plot the angle the pendulum subtends to the vertical over
time. Note that this is not very noisy.

```{r}
s <- extract(samples,permuted=FALSE)
plot(s[1,1,1:100])
```

Now let us suppose that we don't know the value of $g$ and we can only
observe the horizontal displacement.

```{r}
zStan <- sin(s[1,1,1:nSamples])
```

Now we can use Stan to infer the value of $g$.

~~~~{.CPP include="PendulumInfer.stan"}
~~~~

Here's the R to invoke it.

```{r}
estimates <- stan(file = 'PendulumInfer.stan',
                  data = list (
                      T  = nSamples,
                      y0 = m0,
                      z  = zStan,
                      t0 = t0,
                      ts = ts
                  ),
                  seed = 42,
                  chains = 1,
                  iter = 1000,
                  warmup = 500,
                  refresh = -1
                  )
e <- extract(estimates,pars=c("theta[1]","sigma[1]","lp__"),permuted=TRUE)
```

This gives an estiamted valeu for $g$ of `r e$theta[1]` which is what
we would hope.

Now let's try adding some noise to our observations.

```{r}
set.seed(42)
epsilons <- rmvnorm(n=nSamples,mean=c(0.0),sigma=bigR)

zStanNoisy <- sin(s[1,1,1:nSamples] + epsilons[,1])

estimatesNoisy <- stan(file = 'PendulumInfer.stan',
                       data = list (T  = nSamples,
                                    y0 = m0,
                                    z  = zStanNoisy,
                                    t0 = t0,
                                    ts = ts
                                    ),
                       seed = 42,
                       chains = 1,
                       iter = 1000,
                       warmup = 500,
                       refresh = -1
                       )
eNoisy <- extract(estimatesNoisy,pars=c("theta[1]","sigma[1]","lp__"),permuted=TRUE)
```

This gives an estiamted valeu for $g$ of `r eNoisy$theta[1]` which is
ok but not great.

## Postamble

To build this page, download the relevant files from
[github](https://github.com/idontgetoutmuch/RandomWalks) and run this
in R:

    library(knitr)
    knit('Pendulum.Rmd')

And this from command line:

    pandoc -s Pendulum.md --filter=./Include > PendulumExpanded.html


